First paper - Draft

### 2017-06-22 paper/paper.Rnw
Finished reviewing research and data sections

With this being said, this paper introduces one novelty in the literature which is to evaluate the evolution of the high/low SES achievement gap in the past 15 years for all PISA participant countries. This is different from previous work because it concentrates solely on 15 year old children, and it attempts to capture the evolution of the achievement gap for each developed countries. The advantages of this study are numerous. First, we concentrate on the evolution of the gap for only 15 year olds, which will serve as a comparison to the single year-country snapshot of \citet{anna2016} and the evolution of the Kindergarten gap in \citet{reardon_portilla}. As we've seen before, there are reasons to think that specific age-groups have seen changes in the achievement gap. Moreover, in almost all countries with a tracked curriculum children are either at or in the process of tracking by the age 15, meaning that we will be able to link whether tracked countries are the most variable in their evolution of achievement gaps.

@@

This paper is interested in two questions. As explained in the literature review, we already know that U.S achievement gaps have been growing over time, and some international evidence suggests that it might be increasing in other countries as well. All of these studies pool all students together from different ages and estimate global trends.

We take a different approach and we want to study age-specific trends in achievement gaps. We're interested in studying a) the size of the achievement in a comparative perspective and its relationship to overall achievement levels, b) which countries are experiencing changes in the achievement gap, c) the rate at which each country-gap is widening/narrowing and d) establishing whether the gaps are widening/narrowing because particular groups are getting ahead/behind.

@@

a) We want to test the notion that better performing countries have lower leves of achievement gap. \citep{werfhorst_mijs} emphasize that there is empirical evidence that suggests this. This pattern is not so obvious. For example, countries with high leves of tracking could maximize student performance, specially the high SES students, raising their overall performance and thus raising the national performance score. But if the bottom performers are not gaining at the same rate, then the achievement gap will inevitable grow resulting in a high performing countries with widening achievement gap.

b) The seminal work of \citet{reardon2011} suggests that achievement gaps change, and they do so much quicker that we though. After recording a SES gap increase of about 40\% in only 30 years, \citep{reardon_portilla} stress that they also found a significant decrease in only 15 years of data. This lends credibility that our estimations will show variation over time. As usual, their estimations are only for children going from 4th to 8th grade. We concentrate on age-specific groups, in this case, 15 year olds.

c) We want to compare the percentage change at which the gap widened/narrowed from the first to the last year. This will give us a general idea of the overall change over time, and will allow us to compare our estimates to the actual literature \footnote{Although no study has performed this age-specific achievement gap for comparable tests over such a long time. Our results will serve as comparison for other studies that use age-specific groups, such as 4th graders.}

@@

In order the study the above mentioned questions I will be using the Programme for International Student Assessment (PISA). PISA is a survey carried out every three years that aims to evaluate education systems by testing the skills and knowledge of 15-year-old students. Currrently, PISA has six waves starting in the year 2000 up until 2015, where recently, over half a million students were tested in mathematics, literacy and science in over 70 countries, both developed and developing ones.

PISA collects data through a two-stage stratified sampling design. With the help of official governments, PISA randomly chooses 150 schools in each country, where they then randomly pick thirty 15 year olds to undertake the two hour tests. The sample size for PISA 2000 is \Sexpr{country_rows[1]}, for PISA 2003, \Sexpr{country_rows[2]}, for PISA 2006, \Sexpr{country_rows[3]}, for PISA 2009, \Sexpr{country_rows[4]}, for PISA 2012, \Sexpr{country_rows[5]}, and for PISA 2015 \Sexpr{country_rows[6]}. Together with the subject tests, PISA collects personal information from students, their families and their school environment (including teacher surveys), that serves as relevent background information that can be matched to the students performance. With the recent inclusion of PISA 2015, these six waves make up a time-series analysis of 15 years, enough time to visualize changes in the structure of an educational system.

In order to measure family background PISA includes several variables. Firstly, the ask student's their parent's educational level. Scholars have considered this to be a reliable recall given that we expect fifteen year olds to know their parent's level of education \citep{reardon2011}. This question has been asked in every wave and holds a somewhat similar coding across time, although the first two waves have a different coding. Moreover, another limitation is the fact that parent's education is measured with the ISCED classification, something that has changed over time. For example, until PISA 2009, the ISCED 1997 classification was currently used, but the next wave already used the new ISCED 2011 classification. Both these classification schemes have equivalent look-up tables, yet this requires a detailed inspection of the codings.

Another social background variable PISA collects is the International Socio-Economic Index of Occupational Status (ISEI). This variable captures the economic status of the family, the closest it can without asking for income information. It has been scaled for comparability between waves and some authors have used it for inequality studies finding expected results to be consistent with social origin literature \citep{anna2016_global}. PISA also includes a plethora of indicators on family wealth, home educational resources, the number of books in the home, among many other material resources in the household.

But one of the most relevant variables for our study is the is a composite SES index created by the PISA team. The index of economic, social and cultural status (ESCS) was created on the basis of the following variables: the International Socio-Economic Index of Occupational Status (ISEI); the highest level of education of the student’s parents, converted into years of schooling; the PISA index of family wealth; the PISA index of home educational resources; and the PISA index of possessions related to "classical" culture in the family home \citep{oecd_glance_2002}. This variable, aside from capturing all relevant dimensions of SES, such as education, occupation, and material resources, takes care of transforming the variables into comparable measures across waves. In fact, we use the newly-released ESCS index \citep{pisa_2015_results} which rescaled all ESCS indexes to make them suitable for over-time analysis \footnote{These rescaled indices can be found in the \href{http://www.oecd.org/pisa/data/2015database/}{PISA website} under \emph{Rescaled Indices for Trend Analyses}.}.

Aside from SES, the other most relevant variables are test scores. PISA provides plausible values instead of one single indicator of test score performance. As explained in the PISA manual \citep{pisa2012_technical}, these are imputed values that resemble individual test scores and have approximately the same distribution as the latent trait being measured. Suppose we have \(\mu_i\), the average student test score in mathematics for student \(i\). Instead of estimating \(\mu_i\) alone, plausible values estimate a distribution of possible \(\mu\text{'s}\), for student \(i\), together with the likelihood of each \(\mu_i\) based on the respondents answers on the test. This is defined as the posterior distributions of \(\mu\text{'s}\) for student \(i\). The reason why we use this procedure is because estimating a single estimate \(\mu_i\) is plagued with measurement error, among other types of bias \citep[see][]{wu2005}. The plausible values in the PIAAC data are effectively five (ten for PISA 2015) random draws from this distribution. These variable is are continuous, ranging from 0 to 500, with a mean of 250.

@@

The aim of this paper is to identify and disaggregate country trends in the achievement gap for several countries. To represent the SES gap, most of the literature on achievement gaps has concentrated on indicators such as parental education, parental occupational status, income achievement gaps and actual SES achievement gaps \citep{fryer2004}. The actual calculation of the achievement gap varies substantially and different strategies have been implemented. For example, \citet{micklewright} calculate the difference in achievement by crudely subtracting the gap between the 95th and 5th percentile of the mathematics distribution. Although in principle you should be able to capture some type of SES effect like this, theoretically, it should be much more accurate to difference out the mean score of, for example, parental education or some other SES proxy. \citet{saw2016}, for instance, used parental education as a proxy of SES, whereas \citet{byun2010} use a similar SES index as ours, but created by them.

\citet{reardon_portilla, anna2016, anna2016_global} used a different method developed by \citet{reardon2011} which we partially adopt adopt in this paper. SES achievement gaps are measured as the difference in standardized achievement between the 90th and 10th percentiles of the chosen SES variable. The rule of thumb to choose the 90th, 50th and 10th percentile is arbitrary, others have used, for example, the 5th, 50th and 95th \citep{micklewright} while the literature has been following the standard set by \citet{reardon2011} to use 90th/10th percentiles.

We start by standardizing achievement for each country, year and subject. Because PISA does not provide a single achievement indicator, we take the median of all plausible values for each student \footnote{Since each plausible value is a random draw from a theoretical latent normal distribution of possible student achievement scores, the median should be precise in getting a central measure of the latent distribution.}, resulting in one single score.

We then fit a linear model

@@

Standardizing the median test score solves the problem of comparability of gaps measured with different tests, and across waves. However, if the variance of academic achievement changes over time, then standardizing the overall score at each wave of testing confounds changes in the gap with changes in the variance of test scores. That is, by standardizing we're forcing the variance of test scores to be zero across all waves. But if the true variance of academic achievement grows over time, then the estimated trend in the achievement gaps will be underestimated, and vice versa.

% Show graph of variance over time
% I'm showing standard deviations, is it better to show variance?
<<variance_pisa, cache = TRUE>>=

@@

Inspection of the data suggests that it's something we shouldn't be deeply concerned with. The standard deviation of each wave seems to be following a very similar pattern with not so drastic exception of the year 2000.

@@

where \begin{math}r\end{math} is the reliability score of the wave. \footnote{Other proceedures multiply each country by their own reliability measure for each year-subject pair (Anna analysis of wordwide gap). Unfortunately, PISA 2000 did not provide any reliability measure separately for each country and PISA 2015 has to yet release their own. At the moment of writing this paper, they were unavailable. For this reason we implement the analysis following the original work of \citet{reardon2011}}. Each PISA survey provides a reliability indicator which we use accordingly. This yields estimates of the true gaps, and eliminates any bias in the trend that may arise from differential reliability of the tests.

After constructing the adjusted test score measurement, we estimate the SES dummy by calculating the weighted 90th and 10th percentile scores for the SES composite index, and then generating a dummy of 1 for those above (including) the 90th percentile and 0 for those below (including) the 10th percentile.

@@

where \(SES_i\) is whether the student is in the top/bottom SES dummy and we allow both \(\alpha_i\) and \(\beta_i\) to vary by country \(j\). We implement this separately for each wave and weight by the student sample weights.

>>>>

@@

With this being said, this paper introduces one novelty in the literature which is to evaluate the evolution of the high/low SES achievement gap in the past 15 years for all PISA participant countries. This is different from previous work because it concentrates solely on 15 year old children, and it attempts to capture the evolution of the achievement gap for many countries. The advantages of this study are twofold.

First, we concentrate on the evolution of the gap for only 15 year olds, which will serve as a comparison to the single year-country snapshot of \citet{anna2016} and the evolution of the Kindergarten gap in \citet{reardon_portilla}. As we've seen before, there are reasons to think that specific age-groups have seen changes in the achievement gap. Moreover, in almost all countries with a tracked curriculum children are either at or in the process of tracking by the age 15, meaning that we will be able to link whether tracked countries are the most variable in their evolution of achievement gaps.

Second, we will be able, for the first time, to study the evolution of the achievement gap for several countries other than the United States. This will help asses the mangitude of the inequality as well as the rate of growth/decline. Thirdly, this paper not only documents the size and changes of the SES gap, but studies with particular attention the source of the changes. That is, we study whether bottom performers are falling behind, if top performers are gaining advantages, or if both phenomenas are simultaneously at play.

More formally, We're interested in studying
a) The size of the achievement in a comparative perspective and its relationship to overall achievement levels;
b) Which countries are experiencing changes in the achievement gap;
c) the rate at which each country-gap is widening/narrowing and;
d) establishing whether the gaps are widening/narrowing because particular groups are getting ahead/behind.

@@

a) We want to test the notion that better performing countries have a smaller degree of inequality than other countries. \citep{werfhorst_mijs} emphasize that there is empirical evidence that suggests this. This pattern is not so obvious, however. For example, countries with high leves of tracking could maximize student performance, specially the high SES students, raising their overall performance and thus raising the national performance score. But if the bottom performers are not gaining at the same rate, then the achievement gap will inevitable grow resulting in a high performing countries with a widening achievement gap.

b) The seminal work of \citet{reardon2011} suggests that achievement gaps change, and they do so much quicker that we though. After recording a SES gap increase of about 40\% in only 30 years, \citep{reardon_portilla} stress that they also found a significant decrease in only 15 years of data. This shows that it is important to study the changes in the achievement gap over time. This will bring a useful comparison and pinpoint which countries are doing the best to reduce the gap.

c) We want to compare the percentage change at which the gap widened/narrowed from the first to the last year available. This will give us a general idea of the overall change over time, and will allow us to compare our estimates to the actual literature \footnote{Although no study has performed this age-specific achievement gap for comparable tests over such a long time. Our results will serve as comparison for other studies that use age-specific groups, such as 4th graders.}

@@

To investigate the above mentioned questions I will be using the Programme for International Student Assessment (PISA). PISA is a survey carried out every three years that aims to evaluate education systems by testing the skills and knowledge of 15-year-old students. Currrently, PISA has six waves starting which started in 2000 up until 2015, where recently, over half a million students were tested in mathematics, literacy and science in over 70 countries, both developed and developing ones.

PISA collects data through a two-stage stratified sampling design. With the help of official governments, PISA randomly chooses 150 schools in each country, where they then randomly pick thirty 15 year olds to undertake the two hour tests. The sample size for PISA 2000 is \Sexpr{country_rows[1]}, \Sexpr{country_rows[2]} for PISA 2003, \Sexpr{country_rows[3]} for PISA 2006, \Sexpr{country_rows[4]} for PISA 2009, \Sexpr{country_rows[5]} for PISA 2012, and \Sexpr{country_rows[6]} for PISA 2015. Together with the subject tests, PISA collects personal information from students, their families and their school environment (including teacher surveys), that serves as relevent background information that can be matched to the students performance. With the recent inclusion of PISA 2015, these six waves make up a time-series analysis of 15 years, enough to visualize changes in the structure of an educational system. None of the literature cited so far has used the last PISA wave, which was recently released in December 2016.

To identify a family's socio-economic status PISA collects several variables that measure different dimensions. Classically, they ask student's their parent's educational level. Scholars have considered this to be a reliable recall given that we expect fifteen year olds to know their parent's level of education \citep{reardon2011}(cite more here). This question has been asked in every wave and holds a somewhat similar coding across time, although the first two waves have a different coding. In spite of this, another limitation is the fact that parent's education is measured with the ISCED classification, something that has changed over time. For example, until PISA 2009, the preferred framework was ISCED 1997, whereas the next wave switched to the newly developed ISCED 2011 classification. Both these classification schemes have equivalent look-up tables, but this requires a detailed inspection of the codings. Aside from all these limitations, this is very reliable and important variable that measures an important factor of socio-economic status.

Another social background variable PISA collects is the International Socio-Economic Index of Occupational Status (ISEI). This variable captures the economic status of the family, the closest it can without asking for income information.  This index variable was developed by Ganzeboom (cite) and it attempts to measure occupational status using a continuous measure. This indicator has been found to ourperform classical occupational classifications such as the Erikson-Goldthorpe-Portocarero schema (cite). It has been scaled for comparability between waves and some authors have used it for inequality studies finding expected results to be consistent with social origin literature \citep{anna2016_global}. PISA also includes a plethora of indicators on family wealth, home educational resources, the number of books in the home, among many other material resources in the household.

Yet one of the most relevant variables for our study is a composite SES index created by the PISA team. The index of economic, social and cultural status (ESCS) was created on the basis of the following variables: the International Socio-Economic Index of Occupational Status (ISEI), the highest level of education of the student’s parents, the PISA index of family wealth (which measured the material wealth of the family), the PISA index of home educational resources; and the PISA index of possessions related to "classical" culture in the family home (mainly about books in the household) \citep{oecd_glance_2002}. This variable, aside from capturing all relevant dimensions of SES, such as education, occupation, and material resources, takes care of transforming the variables into comparable metrics across waves. To the best of our knowledge, this is the first paper that uses the newly-released ESCS index \citep{pisa_2015_results} which was rescaled so that all ESCS indexes are suitable for over-time analysis \footnote{These rescaled indices can be found in the \href{http://www.oecd.org/pisa/data/2015database/}{PISA website} under \emph{Rescaled Indices for Trend Analyses}.} In other words, the ESCS index does not need any transformation or coding updates because it is ready for comparison over time.

Aside from SES, the other most relevant variables are test scores for Mathematics, Literacy and Science. PISA does not provide a single test result for each respondent. Instead, it provides a \emph{series} of 'plausible values' that the child could actually score. As explained in the PISA manual \citep{pisa2012_technical}, these are imputed values that resemble individual test scores and have approximately the same distribution as the latent trait being measured (the true distribution of the possible scores a student can achieve). Suppose we have \(\mu_i\), the average student test score in Mathematics for student \(i\). Instead of estimating \(\mu_i\) alone, plausible values estimate a distribution of possible \(\mu\text{'s}\) for student \(i\), together with the likelihood of each \(\mu_i\) based on the respondents answers on the test. This is defined as the posterior distributions of \(\mu\text{'s}\) for student \(i\). The reason why we use this procedure is because estimating a single estimate \(\mu_i\) is plagued with measurement error, among other types of bias \citep[see][]{wu2005}. The number of plausible values for PISA waves are usually five (although ten for PISA 2015) random draws from this distribution. In practice each student has 5 scores for each test, that resembles this distribution. They are continuous, ranging from 0 to 500, with a mean of 250.

@@

The aim of this paper is to identify and disaggregate country trends in the achievement gap for several countries. To represent the SES gap, most of the literature on achievement gaps has concentrated on indicators such as parental education, parental occupational status, income achievement gaps and actual SES achievement gaps \citep{fryer2004, hanushek_woesmann_tracking, saw2016, bradbury2015, byun2010}. The actual calculation of the achievement gap varies substantially and different strategies have been implemented. For example, \citet{micklewright} calculates the difference in achievement by crudely subtracting the gap between the 95th and 5th percentile of the mathematics distribution. Although in principle you should be able to capture some type of SES effect like this, theoretically, it should be much more accurate to difference out the mean score of, for example, parental education or some other SES proxy. \citet{saw2016}, for instance, used parental education as a proxy of SES, whereas \citet{byun2010} use a similar SES index as ours, but created by them.

\citet{reardon_portilla, anna2016, anna2016_global} used a different method developed by \citet{reardon2011} which we partially adopt in this paper. SES achievement gaps are measured as the difference in standardized achievement between the 90th and 10th percentiles of the chosen SES variable. The rule of thumb to choose the 90th, 50th and 10th percentile is arbitrary, others have used, for example, the 5th, 50th and 95th \citep{micklewright} while the literature has been following the standard set by \citet{reardon2011} to use 90th/10th percentiles.

For each country in each study, SES disparities in achievement are measured as the gap in standardized achievement between the 90th and 10th percentiles of each country’s distribution of each SES variable, following Reardon’s (2011) method for income achievement gaps.

The original strategy of \citet{reardon2011} is as follows: first, achievement is standardized (see below for a mathematical explanation of the standardization). We then use it to calculate the mean achievement (and standard error) for each category of the SES variable of interest (parent's education, income categories, etc..). "Category means are plotted at their percentile ranks and cubic models are fit through the points using weighted least squares. Finally, achievement at each country’s 90th and 10th SES percentiles is interpolated from the model" \citep{anna2016_global}. The result is a SES gap from an ordinal variable of interest.

As mentioned before, PISA does not provide a single achievement indicator. Instead, we take the median of all plausible values for each student \footnote{Since each plausible value is a random draw from a theoretical latent normal distribution of possible student achievement scores, the median should be precise in getting a central measure of the latent distribution.}, resulting in one single score.

To standardize the test score we fit a linear model

@@

This new standardized variable has a mean of zero. Standardizing the median test score solves the problem of comparability of gaps measured with different tests, and across waves. However, if the variance of academic achievement changes over time, then standardizing the overall score at each country-wave pair actually makes the variable more biased. That is, by standardizing we're forcing the standard deviation of test scores to be zero across all waves. But if the true deviations of the median academic achievement changes over time, then the estimated trend in the SES gaps will be underestimated, or vice versa.

% Show graph of variance over time
% I'm showing standard deviations, is it better to show variance?
<<variance_pisa, cache = TRUE, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=

@@

Inspection of the data suggests that it's something we shouldn't be deeply concerned with. The standard deviation of each wave seems to be following a very similar pattern with a not so drastic exception of the year 2000.

@@

where \begin{math}r\end{math} is the reliability score of the wave. \footnote{Other proceedures multiply each country by their own reliability measure for each year-subject pair \citep{anna2016_global}. Unfortunately, PISA 2000 did not provide any reliability measure separately for each country and PISA 2015 has to yet release their own. At the moment of writing this paper, they were unavailable. For this reason we implement the analysis following the original work of \citet{reardon2011}}. Each PISA survey provides a reliability indicator which we use accordingly. This yields estimates of the true gaps, and eliminates any bias in the trend that may arise from differential reliability of the tests.

After constructing the adjusted test score measurement, we estimate the SES dummy by calculating the weighted 90th and 10th quantiles for the SES composite index, and then generating a dummy of 1 for those above (including) the 90th percentile and 0 for those below (including) the 10th percentile.

@@

where \(SES_i\) is whether the student is in the top/bottom SES dummy and we allow both \(\alpha_i\) and \(\beta_i\) to vary by country \(j\). We implement this separately for each wave and weight by the wave-specific student sample weights.
